---
title: "Cyclistic Bike-Share Analysis"
author: "Onyedikachukwu Okonkwo"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning = FALSE, message = FALSE)
options(repos = c(CRAN = "https://cloud.r-project.org/"))
library(tidyverse)
library(ggplot2)
library(scales)

source("Google_Certificate_Case_Study_1.R")

```
## Executive Summary

This analysis examines usage patterns between casual riders and annual members of Cyclistic bike-share to inform marketing strategies for converting casual riders into annual members.

**Key Findings:**
- Casual riders take longer trips (22.6 min avg) but members ride 12x more frequently
- Members peak on weekdays, casuals peak on weekends
- Casuals take round trips 8.8x more often than members

These patterns suggest casual riders use bikes for leisure/recreation while members use them for daily commuting.

---

## Business Question

**How do annual members and casual riders use Cyclistic bikes differently?**

This analysis uses Q1 2019 and Q1 2020 Divvy trip data (770,000+ rides) to identify behavioral differences between Casuals, and Members that can inform targeted marketing campaigns.

---

## Finding 1: Duration vs. Frequency

Casual riders take significantly longer trips, but members ride much more frequently.
```{r duration-comparison}
ggplot(ride_length_comparison, aes(x = member_casual, y = avg_ride_time, fill = member_casual)) +
  geom_col() +
  labs(
    title = "Average Ride Duration: Casual vs Member",
    x = "Rider Type",
    y = "Average Duration (minutes)",
    fill = "Rider Type"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("casual" = "#FF6B6B", "member" = "#4ECDC4"))

```

**Insight:** Casual riders average 22.6 minutes per trip compared to members' 10.8 minutes, suggesting leisurely rides. However, members take 714,215 total trips versus casuals' 58,316 trips - a 12x difference indicating daily usage patterns.

---

## Finding 2: Weekly Patterns

Members and casuals show opposite weekly patterns.
```{r weekly-patterns}
ggplot(day_of_week_patterns, aes(x = day_name, y = n, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "Ride Patterns Throughout the Week",
    x = "Day of Week",
    y = "Number of Rides",
    fill = "Rider Type"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("casual" = "#FF6B6B", "member" = "#4ECDC4")) +
  scale_x_discrete(limits = day_names)+
  scale_y_continuous(labels = scales::label_number( scale = 1/1000, suffix = "K"))

```

**Insight:** Members peak on weekdays (Tuesday: 127K rides) and drop on weekends, consistent with commuting behavior. Casuals peak on weekends (Sunday: 15K rides) and stay low during weekdays, consistent with recreational use.

---

## Finding 3: Round Trip Behavior

Casuals are far more likely to return to their starting station.
```{r round-trips}
round_trip_viz <- round_trip_pattern %>%
  filter(round_trip == TRUE)

ggplot(round_trip_viz, aes(x = member_casual, y = percentage, fill = member_casual)) +
  geom_col() +
  labs(
    title = "Percentage of Trips That Are Round Trips",
    x = "Rider Type",
    y = "Percentage of Total Rides",
    fill = "Rider Type"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("casual" = "#FF6B6B", "member" = "#4ECDC4")) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), vjust = -0.5)

```

**Insight:** 12.1% of casual trips are round trips compared to only 1.4% of member trips. This supports the leisure vs. commute hypothesis - casuals explore and return, members travel point-to-point.

---

## Finding 4: Total Usage

Members dominate total ridership.
```{r total-rides}
ggplot(ride_length_comparison, aes(x= member_casual , y = total_rides , fill = member_casual)) +
  geom_col() +
  labs(
    title = "Total Number of Rides: Casual vs Member",
    x = "Rider Type",
    y = "Total Rides",
    fill = "Rider Type"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("casual" = "#FF6B6B", "member" = "#4ECDC4")) +
  scale_y_continuous(labels = scales::label_number(scale = 1/1000, suffix = "K")) +
  geom_text(aes(label = scales::comma(total_rides)), vjust = -0.5)
```

**Insight:** With 714,215 trips versus 58,316, members account for 92% of total rides, making them the primary user base despite having shorter individual trip durations.

---

## Recommendations

Based on these behavioral differences, marketing strategies to convert casual riders should consider:

1. **Value Proposition Challenge:** Casuals use bikes for occasional leisure, not daily commuting. Traditional annual memberships may not appeal to this usage pattern.

2. **Potential Strategies:**
   - Create weekend/seasonal pass options tailored to leisure riders
   - Develop campaigns showing how bikes can replace short commutes
   - Offer trial periods that encourage behavioral change toward weekday usage

3. **Target Messaging:** Emphasize flexibility, cost savings for weekend warriors, or lifestyle benefits of incorporating bikes into daily routines.

---

## Methodology

**Data:** Divvy Trips Q1 2019 & Q1 2020 (combined dataset)  
**Tools:** R, tidyverse, ggplot2  
**Process:** Data cleaning, outlier removal (rides <1 min or >60 min), comparative analysis  
**Sample Size:** 772,531 rides analyzed

---
